Stochastic blockmodeling of relational event dynamics

### 2012-10-31 docs/submission.tex
Make novelty a bit more clear.

One can extend the above to marked point processes where each event contains additional information.
In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} comprised of all allowed dyads.
We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{ij}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.  
%The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
The likelihood of an observed event history $\mathcal{A}_{t_M}$ (extending from time 0 to the time of the final event and denoted $\mathcal{A}$ for convenience) is
%The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{ij}(t_m | \cdot) \}
\label{eqn:llk}
\end{align}

@@

Here, we aim to learn about the dynamics within and between subsets of nodes.   To facilitate this, we assume each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{ij}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{N}_P(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.

% In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{ij}(t)$ to change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.

@@

% TODO: Awkward next sentence
The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
% TODO: Also, need to be restricted to ego conditions if we want it fast

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{ij}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

@@

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]} \ \   \times \\
&\prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = K+1 |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} \ \times \\
& \ \ \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{ij}(t_m|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.  We sample each $\beta_{k,l,p}$ via slice sampling.

@@

@@

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 groups where 1) within group communication is more likely, 2) events among members in the first group are more likely to be reciprocated  (i.e. a positive \texttt{ab-ba} effect), and 3) events among members of the second group are more likely to be followed by an event with the same sender (i.e. a positive \texttt{ab-ay} effect).  The specification of $\textbf{s}$ is therefore $\textbf{s}(t,i,j,\mathcal{A}_t) = [s_0, s_{1}(t,i,j,\mathcal{A}_t), s_{2}(t,i,j,\mathcal{A}_t)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
Data is generated by sequentially computing $\lambda_{ij}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{ij} \lambda_{ij}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{ij}(t_m|\cdot) / \sum_{ij}\lambda_{ij}(t_m|\cdot))$.

@@

A variety of real world data sets are used to explore the efficacy of the model.  The following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.  For each data set, we hold out the final $M_{test}$ events for evaluation of the model.
\begin{list}{\labelitemi}{\leftmargin=1em}
%\begin{itemize}
%\item Kiel email: \cite{Ebel2002}
\item Classroom \cite{McFarland2001}: 445 directed communication among 27 people a high
  school classroom collected via participant observation ($M_{test}= 145$).
\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88
  users with at least 30 emails ($M_{test} = 1300$).
\item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
  individuals between July 2001 and August 2001 ($M_{test} = 1000$).
%\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
\item Twitter direct messages: Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This dataset will be made publicly available.}
\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the
  89 recipients between October 2001 and February 2002 ($M_{test}= 1000$).
%\end{itemize}
\end{list}

@@

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.
We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
For this example we fit a model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.
% The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
%The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
% Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
 block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data, and furthermore our inferences from the parameter estimates are sensible given the observed statistics.
{\color{red} TODO: Elaborate on what this figure shows.}

@@

Several baselines are included for comparison: \texttt{uniform} places uniform probability on all possible dyads, \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i,j_m=j)$, and \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i) \sum_{m:t_m < t} \mathbf{I}(j_m=j)$.  
Finally, \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} and \texttt{BM} can capture individual heterogeneity and group level heterogeneity in overall activity.

@@

We additionally perform experiments that evaluate the likelihood of future data under our model.   As our method jointly models \emph{which} dyads occur and \emph{when} they occur, any alternative method needs to also model these two aspects.  In this vein, we extend the above baselines by assuming each dyad is a Poisson process with estimated  rate $$\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{ij} r_{b}(i,j) + \xi}$$, where $r_b(i,j)$ is the statistic for a given baseline (described above) and $\xi=1$ is a smoothing parameter.  
In Table ~\ref{tab:results} we compare the loglikelihood of held-out test data under our proposed model to each of these baselines.  In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
Note that the model with $K*=1$ is an important baseline -- it is simply a relational event model with no groups. There are a large number of other possible baselines that one might use, but most are not directly applicable to Table 2 as they do not work with events in continuous time.

@@

%\subsubsection*{Acknowledgements}
\newpage
\newpage
\bibliographystyle{unsrt}
\bibliography{refs}
\newpage
\textbf{Supplementary material}
\input{appendices}
\end{document}

>>>>

One can extend the above to marked point processes where each event contains additional information.
In the context of relational events occurring among $N$ nodes, each event in the process contains both a sender $i$ and a recipient $j$ such that  $(i,j) \in \mathcal{R}$, where $\mathcal{R}$ is the \emph{risk set} is defined as the set of all possible dyadic events.
We assume each event involving dyad $(i,j) \in \mathcal{R}$ occurs as a Poisson process with intensity $\lambda_{i,j}(t|\cdot)$ that is piecewise constant and depends on the previous history of events  $\mathcal{A}_t = \{(t_m,i_m,j_m): t_m \in [0,t) \}$.  
%The likelihood of an observed event history $\mathcal{A}$ extending from time 0 to the time of the final event, $t_M$, is 
The likelihood of an observed event history $\mathcal{A}_{t_M}$ (extending from time 0 to the time of the final event and denoted $\mathcal{A}$ for convenience) is
%The likelihood of an observed event history $\mathcal{A}_{t_M}$ (also denoted $\mathcal{A}$ for convenience) is
\begin{align}
\mathcal{L}(\mathcal{A}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \prod_{(i,j) \in \mathcal{R}}\exp\{ - (t_m - t_{m-1}) \lambda_{i,j}(t_m | \cdot) \}
\label{eqn:llk}
\end{align}

@@

Here, we aim to learn about the dynamics within and between subsets of nodes.   To facilitate this, we propose a novel extension which assumes each node $i$ belongs to latent cluster $z_i$ and use a log linear model for the intensity functions
\begin{align*}
\log \lambda_{i,j}(t | \mathcal{A}_t,\mathbf{\beta},\mathbf{z}) &= \boldsymbol{\beta}'_{z_i,z_j} \mathbf{s}(t,i,j,\mathcal{A}_t)
%z_i &\sim \mbox{CRP}(\alpha) &
%\boldsymbol{\beta}_{k,l} &\sim %\mbox{N}_p(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)
\end{align*}
where for each pair of clusters $(k,l)$ we have a parameter vector $\boldsymbol{\beta}_{k,l} \sim \mbox{Normal}(\boldsymbol{\mu},\boldsymbol{\sigma}^2I)$ that corresponds to the $P$-dimensional vector of statistics $\mathbf{s}(t,i,j,\mathcal{A}_t)$ computed from the previous history $\mathcal{A}_t$.
Thus, the rate of $(i,j)$ events has the same parameters as other events occurring between group $z_i$ and $z_j$.

% In practice we may believe an interaction between one dyad should not affect the rate of an entirely separate dyad.
As shown in Figure \ref{fig:example}, we only allow each intensity function $\lambda_{i,j}(t)$ to change following an event where $i$ was the sender or the recipient.
This is sensible in distributed settings where $i$ has limited knowledge about interactions among other actors.
In the Supplementary Material we show how this assumption also reduces the computational complexity of computing the above likelihood.

@@

{\color{red} TODO: Awkward next sentence...}
The next set of statistics are \emph{participation shift} effects inspired by research in conversational norms \cite{Gibson2003}.
For example, an \texttt{ab-ba} effect indicates an increased propensity for reciprocity where the event $(a,b)$ is followed by $(b,a)$.
Though only the statistics in Table \ref{tab:stats} are used in our experiments, one may use any quantity that is computed using the previous history of events or known covariates about nodes or dyads.
The only restriction is that the statistic may not change in value between each observed event.%\footnote{This prevents one from using the number of events occurring in some previous time window, as in \cite{Gunawardana2011}.}
% TODO: Also, need to be restricted to ego conditions if we want it fast

@@

Our formulation is reminiscent of the stochastic blockmodel \cite{Nowicki2001} for static networks which models the probability of a dyad as $p(y_{i,j}) =\mbox{logit}^{-1}( \eta_{z_i,z_j})$ where $\eta_{z_i,z_j}$ is interpreted as a mixing rate between group $z_i$ and group $z_j$.
In the proposed method, however, the blockmodel structure facilitates the study of intra-group and inter-group dynamics via a continuous-time network model.

@@

We use Gibbs sampling to sample the latent class assignments $\mathbf{z}$ from the conditional distribution
\begin{align*}
p(z_r | \mathbf{z}_{-r},\mathcal{A},\alpha,\boldsymbol{\beta}) \propto&  p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) p(z_r | \mathbf{z}_{-r},\alpha) \\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) \propto &
\prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[r \in \{i_m,j_m\}]} \ \   \times \\
&\prod_{(i,j) \in \mathcal{U}_r} \exp \{ -(t_m - t_{m-1}) \lambda_{i,j}(t_m|\cdot)\}
\end{align*}
where $\mathcal{U}_r = \{(i,j) \in \mathcal{R}: r \in \{i,j\}\}$ is the set of dyads involving node $r$.
Under a CRP($\alpha$) prior, we have $p(z_i = k | z_{-i},\alpha) = n_k $ and $p(z_i = K+1 |  z_{-i},\alpha) = \alpha$ where $n_k$ is the number of nodes assigned to cluster $k$.

@@

For each block $(k,l)$ we need to sample the vector of parameters $\boldsymbol{\beta}_{k,l}$ from its posterior
\begin{align*}
p(\boldsymbol{\beta}_{k,l} | \mathcal{A}, \textbf{z}, \boldsymbol{\mu}, \boldsymbol{\sigma}) &\propto p(\boldsymbol{\beta}_{k,l} | \boldsymbol{\mu}, \boldsymbol{\sigma}) p( \mathcal{A}| \textbf{z}, \boldsymbol{\beta}) \\
p(\boldsymbol{\beta}_{k,l} | \mu, \sigma) &= \prod_{p=1}^Pp(\beta_{k,l,p}|\mu_p,\sigma_p^2)\\
p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta}) &\propto \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot)^{\mathbf{1}[(i_m,j_m) \in \mathcal{V}_{k,l}]} \ \times \\
& \ \ \prod_{(i,j) \in \mathcal{V}_{k,l}} \exp \{ -(t_m - t_{m-1}) \lambda_{i,j}(t_m|\cdot)\}
\end{align*}
where $\mathcal{V}_{k,l} = \{(i,j): z_{i} \in \{k,l\} \ \mbox{or} \ z_{j} \in \{k,l\} \}$ is the set of dyads with a sender in group $k$ and a recipient in $l$.  We sample each $\beta_{k,l,p}$ via slice sampling.

@@

\subsection{Scalability}

Likelihood computation essentially depends on the number of changepoints (i.e. knots) in all of the intensity functions being modeled.   
We take advantage of our restriction on the types of statistics $\mathbf{s}$ to reduce the computational complexity of computing our likelihood  as
\begin{align*}
\mathcal{L}(\mathcal{A}_{t_M}|\theta) &= \prod_{m=1}^M \lambda_{i_m,j_m}(t_m|\cdot) \times \\
&  \prod_{(i,j) \in \mathcal{R}_{i_m,j_m}}\exp\{ - (t_m - \tau_{m,i,j}) \lambda_{i,j}(t_m | \cdot) \}
\end{align*}
\noindent where event $m$ is the dyad $(i_m,j_m)$, $\tau_{m,i,j}$ is the time  of the changepoint for $\lambda_{i,j}(t|\cdot)$ prior to the $m$th event, and $\mathcal{R}_{i,j}$ is the set of dyads whose intensity changes if $(i,j)$ occurs.%  \cite{Butts2008}. %\footnote{We further assume 1) all intensity functions change at times $t=0$ and $t=t_M$, and 2) the first event is drawn uniformly from the risk set.}

 By limiting the number changepoints, computing the likelihood  $p(\mathcal{A}|\mathbf{z},\boldsymbol{\beta})$ for Gibbs sampling $z_r$ is $O(|\mathcal{U}_r| \cdot P \cdot N)$, avoiding a factor of $N^2$.  In practice, we precompute $\tau_{m,i,j}$ and $\mathbf{s}(t_m,i,j,\mathcal{A}_t)$ for all $m$, $i$, and $j$.  This can be done in one pass through the data set.

% The Supplementary Material mentions one attempt at speeding this up, for larger graphs, but additional assumptions/approximations are likely needed for scalability.

@@

We check our model fitting procedure using a small synthetic data set involving 10 nodes from 2 groups where 1) within group communication is more likely, 2) events among members in the first group are more likely to be reciprocated  (i.e. a positive \texttt{ab-ba} effect), and 3) events among members of the second group are more likely to be followed by an event with the same sender (i.e. a positive \texttt{ab-ay} effect).  The specification of $\textbf{s}$ is therefore $\textbf{s}(t,i,j,\mathcal{A}_t) = [s_0, s_{1}(t,i,j,\mathcal{A}_t), s_{2}(t,i,j,\mathcal{A}_t)]$.  For the synthetic data set we use parameter vectors $\boldsymbol{\beta}_{1,1} = (0,3,0)$,  $\boldsymbol{\beta}_{1,2} = \boldsymbol{\beta}_{2,1} = (-1,0,0)$, and $\boldsymbol{\beta}_{2,2} = (0,2,0)$.
Data is generated by sequentially computing $\lambda_{i,j}(t_m|\cdot)$ for all $(i,j) \in \mathcal{R}$, drawing $t_{m+1}-t_m \sim \mbox{Exp}(\sum_{i,j} \lambda_{i,j}(t_m|\cdot))$, and drawing the dyad $(i,j) \sim \mbox{Categorical}(\lambda_{i,j}(t_m|\cdot) / \sum_{i,j}\lambda_{i,j}(t_m|\cdot))$.

@@

A variety of real world data sets are used to explore the efficacy of the model.  The following data sets are sequences of dyadic events, where each event has a sender, recipient, and timestamp.  For each data set, we hold out the final $M_{test}$ events for evaluation of the model.
\begin{list}{\labelitemi}{\leftmargin=1em}
%\begin{itemize}
%\item Kiel email: \cite{Ebel2002}
\item Classroom \cite{McFarland2001}: 445 directed communication among 27 people a high
  school classroom collected via participant observation ($M_{test}= 145$).
\item University email \cite{Eckmann2004}: 3300 dyadic emails among 88
  users with at least 30 emails ($M_{test} = 1300$).
\item Enron email \cite{Klimt2004}: 4000 dyadic emails among 141
  individuals between July 2001 and August 2001 ($M_{test} = 1000$).
%\item Irvine: dyadic interactions among 401 actors each having more than 30 events,  \cite{Opsahl}
\item Twitter direct messages: Tweets from Twitter.com occurring between from May 11, 2009 to January 26, 2012 that contained the hashtag \texttt{\#rstats}.  This hashtag is used to denote messages pertaining to the R statistical computing environment and sometimes statistical discussion more generally.  We collect dyadic events by selecting tweets beginning with the \texttt{@} symbol (called a \emph{mention}), and mark the first mentioned user as the recipient.
Of 28337 total tweets in this time period, 3926 were directed events among a total of 1079 users.
We use a subset 4330 events among 487 users who participated in more than one event ($M_{test} = 1330$).\footnote{This data set will be made publicly available.}
\item MIT Reality Mining \cite{Eagle2009}: 2000 phone calls among the
  89 recipients between October 2001 and February 2002 ($M_{test}= 1000$).
%\end{itemize}
\end{list}

@@

Figure \ref{fig:parmats} uses a fitted model to provide an example of the differences in the dynamics that can exist between two groups of dyads with similar rate of occurrence.  In this particular example we used the University data set and fit the proposed model with an intercept $s_0$, $\texttt{ab-ba}$ effects $s_1$, and $\texttt{ab-by}$ effects $s_3$.

We focus on the interactions between members of block 1 to members of block 3; 179 events occurred originating from a member of group 3 and 178 originated from group 1.
% The two panels on the left show the variation in $s_0(t_M,i,j,\mathcal{A}_t)$, $s_1(t_M,i,j,\mathcal{A}_t)$ and $s_3(t_M,i,j,\mathcal{A}_t)$ across the dyads $(i,j)$ in blocks $(1,3)$ and $(3,1)$, respectively.
%The boxplots on the right describe the posterior samples of the corresponding parameters $\beta_{1,3,p}$ and $\beta_{3,1,p}$.
% Under the model, events from group 3 to group 1 occur roughly $e^1$ times as often.
 block (3,1) has a higher propensity for \texttt{ab-by} transitions under the model.
The model has learned that this structure exists in the data, and furthermore our inferences from the parameter estimates are sensible given the observed statistics.
{\color{red} TODO: Elaborate on what this figure shows.}

@@

Several baselines are included for comparison: 
\begin{enumerate}
\item \texttt{uniform} places uniform probability on all possible dyads, 
\item \texttt{online} ranks events at time $t$ by the number of times the dyad has occurred previously $r_{online}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i,j_m=j)$, 
\item \texttt{marginal} uses the product of the observed marginal frequencies $r_{marg}(m,i,j) = \sum_{m:t_m < t} \mathbf{I}(i_m=i) \sum_{m:t_m < t} \mathbf{I}(j_m=j)$.  
\item \texttt{BM} is a stochastic blockmodel (i.e. our model with only an intercept term).
\end{enumerate}
Note for processes that are homogeneous over time, \texttt{online} should do well with large amounts of data while \texttt{marginal} and \texttt{BM} can capture individual heterogeneity and group level heterogeneity in overall activity.

@@

We additionally perform experiments that evaluate the likelihood of future data under our model.   As our method jointly models \emph{which} dyads occur and \emph{when} they occur, any alternative method needs to also model these two aspects.  In this vein, we extend the above baselines by assuming each dyad is a Poisson process with estimated  rate $$\hat{\lambda}_{i,j}(t_m) = \frac{M}{t_M} \frac{r_{b}(i,j) + \xi}{\sum_{i,j} r_{b}(i,j) + \xi}$$, where $r_b(i,j)$ is the statistic for a given baseline (described above) and $\xi=1$ is a smoothing parameter.  
In Table ~\ref{tab:results} we compare the loglikelihood of held-out test data under our proposed model to each of these baselines.  In order to investigate the role of the number of clusters, we introduce an upper bound $K^*$ on the number of clusters during the fitting process.
Note that the model with $K*=1$ is an important baseline -- it is simply a relational event model with no groups. There are a large number of other possible baselines that one might use, but most are not directly applicable to Table 2 as they do not work with events in continuous time.

@@

%\subsubsection*{Acknowledgements}
\newpage
\newpage
\bibliographystyle{unsrt}
\bibliography{refs}
\end{document}
